---
title: "Fecal Microbiota, Fecal Metabolome, and Colorectal Cancer Interrelations"
author: "Chien-Hung Lu"
date: "2024-08-17"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)
library(tidyverse)
library(ComplexHeatmap)

# Import csv data
demo <- read.csv("https://raw.githubusercontent.com/VilcekSummerR/final-assignment-Howard790606/main/S4Table.csv")

mt <- read.csv("https://raw.githubusercontent.com/VilcekSummerR/final-assignment-Howard790606/main/S2Table.csv")

mb <- read.csv("https://raw.githubusercontent.com/VilcekSummerR/final-assignment-Howard790606/main/S3Table.csv")

```

## Background

- This dataset is from NYU Data Catalog.

- This study aims to investigate the interactions between fecal microbiota and metabolites, either with colorectal adenocarcinoma (CRC) or not.

- First, we want to have an overview of the dataset. We first check the demographics.

- Link: https://datacatalog.med.nyu.edu/dataset/10121
```{r demo, echo = TRUE}

# Check the first 5 rows of data
head(demo)

# Subsetting the first 8 columns as demographics data
demo <- demo[,c(1:8)]

```

## Data cleaning and refactor the values

```{r split dataset, echo = TRUE}

# Factor and relabel the numeric data
demo$case <- factor(demo$case, labels = c('patient','control'))
demo$sex <- factor(demo$sex, labels = c('male','female'))
demo$race <- factor(demo$race, labels = c('white','black and others'))

# Create a new column for hospital and factorize
demo$hospital <- ifelse(demo$hosp1 == 1, "A", ifelse(demo$hosp2 == 1, "B", "C"))
demo$hospital <- factor(demo$hospital, labels = c('Walter Reed Army Medical Center', 'George Washington University Hospital', 'National Naval Medical Center'))

# Delete the hosp1 and hosp2 columns
demo <- demo[,c(-7,-8)]

# Have a summary of data
summary(demo)

```

Next, we split the dataset into patient and control group.
```{r dataset, echo = TRUE}
# Use filter method to split data, case 0 refers to patient group, case 1 refers to control group
demo_pat <- demo %>% filter(case == "patient")
  
demo_cont <-demo %>% filter(case == "control")
```

## Visulization of data

- Take a look at the histogram of each continuous variable to see if ther are normally distributed.

- The histograms reveal the high possibilities of normal distribution.
```{r hist, echo = TRUE}
par(mfrow=c(1, 2))
hist(demo$age, col = "steelblue")
hist(demo$bmi, col = "steelblue")
```

- Check skewness by QQPlot and QQline.

Most of our data lie on the QQline, indicating that the data is normally distributed.
```{r qqplot, echo = TRUE}
par(mfrow=c(1,2))
##QQPlot for age
qqnorm(demo$age, main = "Normal Q-Q Plot Age")

#QQLine adds a straight line for reference
qqline(demo$age)

##QQPlot for BMI
qqnorm(demo$bmi,main= "Normal Q-Q Plot BMI")

#QQLine adds a straight line for reference
qqline(demo$bmi)
```

## Use Shapiro test to check if age and bmi are normally distributed

- It shows that age is normally distributed. However, bmi is statistically not normally distributed (p<0.05).

```{r shapiro, echo = TRUE}
shapiro.test(demo$age)
shapiro.test(demo$bmi)
```

## Comparing the difference between patients with CRC and control group

- Compare the patient group and control group using Student T tests for continuous variables and Chi's square tests for categorical variables. 

- The results show that age is statistically different between two groups.

```{r significance check, echo = TRUE}
# Compare sex difference between patient and control groups
chisq.test(demo$sex, demo$case)

# Compare sex difference between patient and control groups
chisq.test(demo$race, demo$case)

# Compare hospital difference between patient and control groups
chisq.test(demo$hospital, demo$case)

# Compare age difference between patient and control groups
t.test(demo$age ~ demo$case)

# Compare BMI difference between patient and control groups
t.test(demo$bmi ~ demo$case)
```

## Use left_join to combine demographics, microbiota data, and metabolites data

- Use ggplot2 to check the data of microbiota(since column 8) and metabolites(since column 228)

```{r leftjoin, echo = TRUE}
demo_mb  <- left_join(demo, mb, by = "ID")
demo_mb_mt  <- left_join(demo_mb, mt, by = "ID")

par(mfrow=c(1,2))

pic <- ggplot(demo_mb_mt, aes(x = Root.k__Archaea, y = X_1_11_UNDECANEDICARBOXYLATE, col = as.factor(case)))

pic + geom_point()

pic2 <- ggplot(demo_mb_mt, aes(x = Root.k__Bacteria, y = X_1_2_PROPANEDIOL, col = as.factor(case)))

pic2 + geom_point()
```

## Check the correlation between fecal microbiota-microbiota and microbiota-metabolite in CRC and control groups

- Use heapmap to draw the corrlelation map.

```{r heapmap, echo = TRUE}

# Use filter method to split data to patient group and control group
demo_mb_mt_pat <- demo_mb_mt %>% filter(case == "patient")
demo_mb_mt_cont <-demo_mb_mt %>% filter(case == "control")

demo_mb_mt_pat <- demo_mb_mt_pat[,c(8:757)]
demo_mb_mt_cont <- demo_mb_mt_cont[,c(8:757)]

par(mfrow=c(1,2))
Heatmap(cor(demo_mb_mt_pat), show_column_names = FALSE, show_row_names = FALSE)
Heatmap(cor(demo_mb_mt_cont), show_column_names = FALSE, show_row_names = FALSE)

```

## Use which() function to find out the most significant microbiota and metabolites pair, and see if there is group difference

- Different patterns of significant correlation pairs are seen.

```{r which, echo = TRUE}

h1 <- cor(demo_mb_mt_pat)
h2 <- cor(demo_mb_mt_cont)

h1_which <- which(((1>h1*h1) & (h1*h1>0.9)), arr.ind=T)
h2_which <- which(((1>h2*h2) & (h2*h2>0.9)), arr.ind=T)

# Find out most significant correlation pair in patient group
head(h1_which)

# Find out most significant correlation pair in control group
head(h2_which)

```
 
## Conclusion

- Age data is normally distributed while bmi data is not normally distributed.

- There is significant difference between patients with colorectal carcinoma (CRC) and controls in terms of age.

- Fecal microbiota-microbiota and microbiota-metabolite correlation shows different pattern within the whole group and between CRC and control group.

## Discussion and future direction

- How to generate a multivariate regression model to see if there is significant independent microbiota-metabolite correlation coefficient is warranted.

- A reliable statistical method to address multi-comparison problem is needed in further study.

- Ref: Sinha R, Ahn J, Sampson JN, et al. Fecal Microbiota, Fecal Metabolome, and Colorectal Cancer Interrelations. PLoS One. 2016;11(3):e0152126. Published 2016 Mar 25. doi:10.1371/journal.pone.0152126